---
title: "Total and Indirect Effect Centrality Measures in CT-Med"
format:
  html:
    embed-resources: true
editor: source
author: Ivan Jacob Agaloos Pesigan
date: February 2, 2026
---

## Installing Required Packages

Before proceeding, ensure that the necessary package are installed and loaded.
We use `cTMed` for calculating total, direct, and indirect effects in continuous-time mediation model.

```{r, include = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
pkg <- c("dynr", "expm", "simStateSpace", "cTMed")
for (i in seq_along(pkg)) {
  if (!require(pkg[i], character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg[i], dependencies = TRUE)
  }
}
```

```{r}
library(cTMed)
```

## Extracting Matrices for Use in `cTMed`

We begin by defining the drift matrix `phi` ($\boldsymbol{\Phi}$).

```{r}
phi <- matrix(
  data = c(
    -0.357, 0.771, -0.450,
    0.000, -0.511, 0.729,
    0.000, 0.000, -0.693
  ),
  nrow = 3,
  ncol = 3
)
phi
```

Next, we assign row and column names to `phi` ($\boldsymbol{\Phi}$), which will be used to index the matrix.

```{r}
colnames(phi) <- rownames(phi) <- c("x", "m", "y")
phi
```

We then create a sequence of time intervals `delta_t` ($\Delta t$) ranging from 0.01 to 10, with a total of 1,000 equally spaced values.

```{r}
delta_t <- seq(from = 0.01, to = 10, length.out = 1000)
```

Let's calculate total and indirect effects centrality for the sequence of `delta_t` ($\Delta t$). We plot to visualize the dynamics.

## Total Effect Centrality

```{r}
total_central <- TotalCentral(
  phi = phi,
  delta_t = delta_t
)
plot(total_central)
```

## Indirect Effect Centrality

```{r}
indirect_central <- IndirectCentral(
  phi = phi,
  delta_t = delta_t
)
plot(indirect_central)
```
