---
title: "Uncertainty Quantification in CT-Med (Centrality Measures)"
format:
  html:
    embed-resources: true
editor: source
author: Ivan Jacob Agaloos Pesigan
date: February 2, 2026
---

## Install the cTMed package

```{r, include = FALSE, message = FALSE, warning = FALSE, results = 'hide'}
pkg <- "cTMed"
for (i in seq_along(pkg)) {
  if (!require(pkg[i], character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg[i], dependencies = TRUE)
  }
}
```

```{r}
library(cTMed)
```

## Extracting Matrices for Use in `cTMed`

After fitting, we extract the estimated drift matrix ($\boldsymbol{\Phi}$) and process noise covariance ($\boldsymbol{\Sigma}$), along with their sampling covariance matrices.

```{r}
fn <- "fit-ct-var-dynr.Rds"
fit <- readRDS(fn)
summary(fit)
```

```{r}
coefs <- coef(fit)
vcovs <- vcov(fit)
phi_names <- c(
  "phi_1_1", "phi_2_1", "phi_3_1",
  "phi_1_2", "phi_2_2", "phi_3_2",
  "phi_1_3", "phi_2_3", "phi_3_3"
)
sigma_names <- c(
  "sigma_1_1", "sigma_2_1", "sigma_3_1",
  "sigma_2_1", "sigma_2_2", "sigma_3_2",
  "sigma_3_1", "sigma_3_2", "sigma_3_3"
)
sigma_vech_names <- c(
  "sigma_1_1", "sigma_2_1", "sigma_3_1",
  "sigma_2_2", "sigma_3_2",
  "sigma_3_3"
)
theta_names <- c(
  "phi_1_1", "phi_2_1", "phi_3_1",
  "phi_1_2", "phi_2_2", "phi_3_2",
  "phi_1_3", "phi_2_3", "phi_3_3",
  "sigma_1_1", "sigma_2_1", "sigma_3_1",
  "sigma_2_2", "sigma_3_2",
  "sigma_3_3"
)
phi <- matrix(
  data = coefs[phi_names],
  nrow = 3,
  ncol = 3
)
colnames(phi) <- rownames(phi) <- c("x", "m", "y")
sigma <- matrix(
  data = coefs[sigma_names],
  nrow = 3,
  ncol = 3
)
theta <- coefs[theta_names]
vcov_phi_vec <- vcovs[phi_names, phi_names]
vcov_sigma_vech <- vcovs[sigma_vech_names, sigma_vech_names]
vcov_theta <- vcovs[theta_names, theta_names]
```

### Estimated Drift Matrix with Corresponding Sampling Covariance Matrix

```{r}
phi
vcov_phi_vec
```

### Process Noise Covariance Matrix with Corresponding Sampling Covariance Matrix

```{r}
sigma
vcov_sigma_vech
```

### Estimated Drift Matrix and Process Noise Covariance Matrix with Corresponding Sampling Covariance Matrix

```{r}
theta
vcov_theta
```

```{r}
delta_t <- seq(from = 0.01, to = 10, length.out = 1000)
```

## Total Effect Centrality

### Delta Method

```{r}
delta_total_central <- DeltaTotalCentral(
  phi = phi,
  vcov_phi_vec = vcov_phi_vec,
  delta_t = delta_t
)
```

```{r}
plot(delta_total_central)
```

### Monte Carlo Method

```{r}
mc_total_central <- MCTotalCentral(
  phi = phi,
  vcov_phi_vec = vcov_phi_vec,
  delta_t = delta_t,
  R = 20000L
)
```

```{r}
plot(mc_total_central)
```

## Indirect Effect Centrality

### Delta Method

```{r}
delta_indirect_central <- DeltaIndirectCentral(
  phi = phi,
  vcov_phi_vec = vcov_phi_vec,
  delta_t = delta_t
)
```

```{r}
plot(delta_indirect_central)
```

### Monte Carlo Method

```{r}
mc_indirect_central <- MCIndirectCentral(
  phi = phi,
  vcov_phi_vec = vcov_phi_vec,
  delta_t = delta_t,
  R = 20000L
)
```

```{r}
plot(mc_indirect_central)
```
